---
title: "ENVS 193DS Spring 2026 Final"
author: "Elena Donelson"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
editor: source
---

[Link to GitHub repository](https://github.com/elenadonelson/ENVS-193DS_spring-2026_final.git)

# Setup

```{r}
#| message: false
#| warning: false

# loading packages
library(tidyverse)
library(janitor)
library(gt)
library(here)
library(DHARMa)

# reading in data
nests <- read_csv(here("data", "nest_data_final.csv"))
```

# Problem 1. Research writing

## a.

In part 1, they likely used a logistic regression because they were testing whether distance to the water edge predicts the probability of American Avocet microhabitat use.

In part 2, they likely used a chi-square test of independence because they were testing whether there is a relationship between bird species and habitat type, which are both categorical variables.

## b.

For part 1, the probability of American Avocet microhabitat use changed with distance from the water edge, suggesting that distance to water may influence where American Avocets choose to use habitat.

For part 2, American Avocets, Forster's Terns, and Black-necked Stilts used habitats differently, suggesting that each species may have different habitat preferences within the San Francisco Bay.

## c. 

One additional variable that could influence the probability of American Avocet microhabitat use is water depth (cm), which could be measured at each bird observation location using a meter stick or depth gauge. Water depth likely influences habitat use because American Avocets are wading birds that feed in shallow water, and the paper found that they were commonly associated with wet bare ground and shorelines where water is typically shallower.

## d. 

# Problem 2. Data analysis

## a.

```{r}
# cleaning and wrangling nest data
nests_clean <- nests %>%
  
  # selecting columns of interest
  select(case_control, height_cm, ant_species) %>%
  
  # keeping only the three ant species needed for analysis
  filter(ant_species %in% c("TP", "AB", "RRB")) %>%
  
  # creating a column with full scientific names
  mutate(
    ant_species_full = case_when(
      ant_species == "TP" ~ "C. sjostedti",
      ant_species == "AB" ~ "C. mimosae",
      ant_species == "RRB" ~ "C. nigriceps"
    )
  ) %>%
  
  # reordering ant species levels
  mutate(
    ant_species_full = factor(
      ant_species_full,
      levels = c(
        "C. sjostedti",
        "C. mimosae",
        "C. nigriceps"
      )
    )
  )

# displaying the structure of the cleaned data frame
str(nests_clean)

# displaying 10 random rows from the cleaned data frame
slice_sample(nests_clean, n = 10)
```

## b. 

In this data set, a value of 1 indicates that a bird nest was present in the tree, while a value of 0 indicates that no bird nest was present. Biologically, this means the response variable represents whether or not a tree was used as a nesting site by birds.

## c.

**Tree height:** Tree height is a continuous variable measured in centimeters. As tree height increases, I would expect the probability of nest occurrence to increase because taller trees may provide birds with better nesting sites and more protection from predators. I found this information in the Introduction section where the authors discuss how habitat characteristics can influence nesting behavior.

**Acacia ant species:** Acacia ant species is a categorical variable with three levels: *Crematogaster sjostedti*, *Crematogaster mimosae*, and *Crematogaster nigriceps*. Different ant species may influence nest occurrence because they vary in their behavior and aggressiveness, which could make some trees more attractive or safer for nesting birds than others. I found this information in the Introduction section where the authors describe how species interactions can affect bird nesting behavior.

## d.

| Model Number | Predictors | Interaction Term? | Model Description |
|--------------|------------|------------------|------------------|
| Model 1 | Tree height, ant species | No | Tests whether tree height and ant species influence the probability of nest occurrence. |
| Model 2 | Tree height, ant species | Yes | Tests whether the effect of tree height on nest occurrence differs among ant species.

## e.

```{r}
# model 1: no interaction
model_1 <- glm(
  case_control ~ height_cm + ant_species_full,
  data = nests_clean,
  family = "binomial"
)

# model 2: interaction model
model_2 <- glm(
  case_control ~ height_cm * ant_species_full,
  data = nests_clean,
  family = "binomial"
)
```

## f. 

```{r}
# comparing models using AIC
AIC(model_1, model_2)
```

The best model as determined by Akaike’s Information Criterion (AIC) included tree height and acacia ant species as predictors of nest occurrence. Including an interaction between tree height and acacia ant species did not improve the model, so the simpler model was selected.

## g. 

```{r}
# creating simulated residuals for the selected logistic regression model
model_1_residuals <- simulateResiduals(
  fittedModel = model_1
)

# plotting diagnostic results
plot(model_1_residuals)

# testing for uniformity of residuals
testUniformity(model_1_residuals)

# testing for dispersion
testDispersion(model_1_residuals)

# testing for outliers
testOutliers(model_1_residuals)

```

The DHARMa diagnostic plots and tests did not show any major issues with the model. The residuals looked fairly even, and there was no evidence of overdispersion or outliers, suggesting that the model fit the data well.

## h.

```{r}
# creating prediction data for the selected model
prediction_data <- expand_grid(
  height_cm = seq(
    min(nests_clean$height_cm, na.rm = TRUE),
    max(nests_clean$height_cm, na.rm = TRUE),
    length.out = 100
  ),
  ant_species_full = levels(nests_clean$ant_species_full)
)

# getting model predictions on the logit scale
model_predictions <- predict(
  model_1,
  newdata = prediction_data,
  type = "link",
  se.fit = TRUE
)

# adding predicted probabilities and 95% confidence intervals
prediction_data <- prediction_data %>%
  mutate(
    fit = model_predictions$fit,
    se = model_predictions$se.fit,
    probability = plogis(fit),
    lower_ci = plogis(fit - 1.96 * se),
    upper_ci = plogis(fit + 1.96 * se)
  )

# plotting model predictions with 95% confidence intervals and raw data
ggplot() +
  
  # plotting the underlying data
  geom_jitter(
    data = nests_clean,
    aes(
      x = height_cm,
      y = case_control,
      color = ant_species_full
    ),
    height = 0.05,
    width = 0,
    alpha = 0.4
  ) +
  
  # adding 95% confidence intervals
  geom_ribbon(
    data = prediction_data,
    aes(
      x = height_cm,
      ymin = lower_ci,
      ymax = upper_ci,
      fill = ant_species_full
    ),
    alpha = 0.25
  ) +
  
  # adding model prediction lines
  geom_line(
    data = prediction_data,
    aes(
      x = height_cm,
      y = probability,
      color = ant_species_full
    ),
    linewidth = 1
  ) +
  
  # using non-default colors
  scale_color_manual(
    values = c(
      "C. sjostedti" = "darkgreen",
      "C. mimosae" = "steelblue",
      "C. nigriceps" = "darkorange"
    )
  ) +
  scale_fill_manual(
    values = c(
      "C. sjostedti" = "darkgreen",
      "C. mimosae" = "steelblue",
      "C. nigriceps" = "darkorange"
    )
  ) +
  
  # putting each ant species in a separate panel
  facet_wrap(~ ant_species_full) +
  
  # labeling axes and title
  labs(
    title = "Predicted probability of nest occurrence by tree height and acacia ant species",
    x = "Tree height (cm)",
    y = "Predicted probability of nest occurrence"
  ) +
  
  # removing grey background and gridlines
  theme_classic() +
  
  # removing the legend
  theme(
    legend.position = "none"
  )
```

## i.

**Figure 1. Predicted probability of bird nest occurrence by tree height (cm) and acacia ant species.** Points represent individual trees where nests were either present (1) or absent (0). Solid lines represent predicted probabilities of nest occurrence from the selected logistic regression model, while shaded bands represent 95% confidence intervals around the predictions. Separate panels represent trees occupied by *C. mimosae*, *C. nigriceps*, and *C. sjostedti*. Data source: Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. 2023. *Data, code, and metadata for: Symbiotic acacia ants drive nesting behavior by birds in an African savanna*. Zenodo. https://doi.org/10.5281/zenodo.8373322.



